The complex structure and peculiar internal motions of the planetary nebula NGC 6210

### 2020-06-15 ms.tex
Write about asymmetries in the nebula

\begin{table*}
  \caption{Physical parameters of nebular components}
  \label{tab:summary}
  \begin{tabular}{
    l % Component
    r % Ha flux
    r % Density
    S % Mass
    r % ioniz param
    S % tstart
    S % tstop
    S % Mdot
    r % Vel
    }
    \toprule
    {}          & {\Ha{} flux} & {Ionized Density} & {Ionized Mass}    & {Ionization Parameter} & {\(t_{\text{start}}\)} & {\(t_{\text{end}}\)} & {\(\dot{M}\)} & {\(V\)}\\
    {Component} & {\% of total}& {\si{H.cm^{-3}}} & {\(M_\odot\)} &          & {\SI{1000}{yr}}      & {\SI{1000}{yr}}    & {\si{\msun.yr^{-1}}} & {\si{km.s^{-1}}}\\
    \midrule
    {(1)} & {(2)} & {(3)} & {(4)} & {(5)} & {(6)} & {(7)} & {(8)} & {(9)}\\
    \addlinespace
    Inner shell & 44\% & 4000--5000 & 0.078 & 0.0036--0.0070 & -4 & -1.5 & 3.1e-5 & 40--55\\
    Intermediate shell & 13\% & 700--1000 & 0.135 & 0.0080--0.0140 & -10 & -4 & 2.3e-5& 35--38\\
    Knot complexes & 42\% & 2000--5000 & 0.063 & 0.0004--0.0050 & -8 & -1.5 & 9.7e-6& 24--67\\
    Outer Lobes & 1\% & 160--320 & 0.024 & 0.0086--0.0224 & -5 & -2 & 8e-6 & 40--90\\
    Inner Halo & 0.5\% & 80--160 & 0.030 & 0.0200--0.0340 & -20 & -10 & 3e-6 & 15: \\
    Outer Halo & 0.2\% & 15--40 & 0.042 & 0.0540--0.0760 & -40 & -20 & 2.1e-6 & 7: \\
    \addlinespace
    NW knot & 0.01\% & 6700 & 1e-5 & 0.0006 & -7.1 & -6.7 & 2.5e-8 & 24\\
    SE knot & 0.08\% & 8200 & 6e-5 & 0.0034 & -1.25 & -0.95 & 2e-7 & 54\\
    N knot &  0.02\% & 3000 & 3e-5 & 0.0012 & -3.0 & -2.7 & 1e-7 & 67\\
    \bottomrule
  \end{tabular}
\end{table*}

@@

The masses can be combined with the kinematic ages (column~10 of Table~\ref{tab:3d}) to estimate the mass loss history of the nebula.
This is given in columns~6 to~8 of Table~\ref{tab:summary} and illustrated in Figure~\ref{fig:mass-loss-history}.
For the start and end times of each component, we take the full range of kinematic times of discrete features in the case of the lobes and knot complexes.
In the case of shells, the start time is taken as the kinematic age of material just outside the shell (for instance, the innermost halo material in the case of the intermediate shell).
In all cases, it is implicitly assumed that no significant acceleration or deceleration has taken place,
which may not be accurate, especially for the inner shell that has been shocked due to the action of the fast wind.
In general, the mass loss is seen to increase monotonically over the last \SI{40000}{yr} from about \SI{2e-6}{\msun.yr^{-1}} to about \SI{3e-5}{\msun.yr^{-1}} in the last few thousand years.
Note that the values shown in Figure~\ref{fig:mass-loss-history} are average values over the respective periods,
so the sudden jump that is apparent \SI{10,000}{yr} ago may have been a more gradual transition.
Also, this method can only estimate mass loss before \SI{1500}{yr} ago, which is the kinematic timescale of the inner shell.
Since that time, the mass-loss rate is expected to have declined significantly
due to the central star transitioning to higher effective temperature and smaller radius,
eventually reaching the present-day value of \(\dot{M} = \SI{9.12e-9}{\msun.yr^{-1}}\) at a wind velocity of \(V = \SI{2150}{km.s^{-1}}\) \citep{Herald:2011a}.
This is indicated on the figure by the gray downward pointing arrow.

@@

A significant fraction of planetary nebulae show deviations even from axisymmetry \citep{Soker:2001a}.
In the cases where these deviations are moderate,
then they may be explained by binary interactions in which the stars are on eccentric orbits,
or the presence of hot or cold spots in the AGB star envelope \citep{Soker:2002b}.
Other cases, where the nebular asymmetries occur only at the outer boundary,
are probably due to interaction with the interstellar medium \citep{Ali:2012a}.
However, a significant minority (\(\sim 10\%\)) of nebulae show such large deviations from bipolar symmetry \citep{Bear:2017a}
that it has been proposed that interacting triple systems are responsible for their shaping \citep{Soker:2004b, Glanz:2020a}.
NGC~6210 falls into this category and has been proposed to be the outcome of either a ``tight binary in a wide orbit'' or a ``tight binary merger'' scenario \citep{Soker:2016b}.
However, given the multitude of evolutionary pathways available in hierarchical triple systems \citep{Toonen:2016a, Toonen:2020a},
other scenarios are also possible.
Sub-arc second imaging observations of young planetary nebulae \citep{Sahai:1998a, Sahai:2011a}
and preplanetary nebulae \citep{Sahai:2007a}
show that such asymmetries can appear very early in the evolution of the nebula,
before the central star is hot enough to emit ionizing photons.

@@

In terms of the GISW model, the halo represents the fossil AGB wind.
The wind is originally neutral,
but it is flash-ionized when the intermediate shell becomes optically thin to EUV radiation.
Since the propagation of the ionization front through the halo is always R-type \citep{Kahn:1954a},
the gas experiences no mechanical effect and maintains its original density and velocity profile.

\subsubsection{Asymmetries}
\label{sec:asymmetries}

Various ways in which the nebula is lop-sided:
\begin{enumerate}[1.]
\item The inner shell emission in \oiii{} and \Ha{} shows brightness asymmetries in both position and velocity.
  The redshifted half of the shell is roughly twice as bright as the blueshifted half,
  while both halves show spatial asymmetries along axis~A, but in opposite senses:
  redshifted emission is 2.5 times brighter on the north side,
  while blueshifted emission is 1.5 times brighter on the south side.
  Because the inner shell dominates the total flux, this leads to a north--south asymmetry in the nebula as a whole,
  with the peak brightness region lying 2~arcsec north of the central star. 
\item In the \heii{} line, the inner shell has an opposite asymmetry: it is brighter to the south than to the north.
   Such an asymmetry is hard to explain if the \chem{He^{2+}} zone of the nebula is ionization-bounded \dots
 \item With the low-ionization knots, the principle asymmetry is in velocity,
   with the redshifted knots being brighter by a factor of 2,
   and also 
\end{enumerate}

@@

>>>>

\begin{table*}
  \caption{Physical parameters of nebular components}
  \label{tab:summary}
  \begin{tabular}{
    l % Component
    r % Ha flux
    r % Density
    S % Mass
    r % ioniz param
    S % tstart
    S % tstop
    S % Mdot
    r % Vel
    }
    \toprule
    {}          & {\Ha{} flux} & {Ionized Density} & {Ionized Mass}    & {Ionization Parameter} & {\(t_{\text{start}}\)} & {\(t_{\text{end}}\)} & {\(\dot{M}\)} & {\(V\)}\\
    {Component} & {\% of total}& {\si{H.cm^{-3}}} & {\(M_\odot\)} &          & {\SI{1000}{yr}}      & {\SI{1000}{yr}}    & {\si{\msun.yr^{-1}}} & {\si{km.s^{-1}}}\\
    \midrule
    {(1)} & {(2)} & {(3)} & {(4)} & {(5)} & {(6)} & {(7)} & {(8)} & {(9)}\\
    \addlinespace
    Inner shell & 70\% & 4000--5000 & 0.078 & 0.0036--0.0070 & -4 & -1.5 & 3.1e-5 & 40--55\\
    Intermediate shell & 13\% & 700--1000 & 0.135 & 0.0080--0.0140 & -10 & -4 & 2.3e-5& 35--38\\
    % Knot complexes & 42\% & 2000--5000 & 0.063 & 0.0004--0.0050 & -8 & -1.5 & 9.7e-6& 24--67\\
    Red knots & 11\% & 2000--5000 & 0.038 & 0.0004--0.0020 & -8 & -3.7 & 9.5e-6& 25-40\\
    Blue knots & 4\% & 1500--2000 & 0.025 & 0.0040--0.0170 & -2.7 & -1.2 & 1.2e-5& 40--65\\
    Outer Lobes & 1\% & 160--320 & 0.024 & 0.0086--0.0224 & -5 & -2 & 8e-6 & 40--90\\
    Inner Halo & 0.5\% & 80--160 & 0.030 & 0.0200--0.0340 & -20 & -10 & 3e-6 & 15: \\
    Outer Halo & 0.2\% & 15--40 & 0.042 & 0.0540--0.0760 & -40 & -20 & 2.1e-6 & 7: \\
    \addlinespace
    NW knot & 0.01\% & 6700 & 1e-5 & 0.0006 & -7.1 & -6.7 & 2.5e-8 & 24\\
    SE knot & 0.08\% & 8200 & 6e-5 & 0.0034 & -1.25 & -0.95 & 2e-7 & 54\\
    N knot &  0.02\% & 3000 & 3e-5 & 0.0012 & -3.0 & -2.7 & 1e-7 & 67\\
    \bottomrule
  \end{tabular}
\end{table*}

@@

The masses can be combined with the kinematic ages (column~10 of Table~\ref{tab:3d}) to estimate the mass loss history of the nebula.
This is given in columns~6 to~8 of Table~\ref{tab:summary} and illustrated in Figure~\ref{fig:mass-loss-history}.
For the start and end times of each component, we take the full range of kinematic times of discrete features in the case of the lobes and knot complexes.
In the case of shells, the start time is taken as the kinematic age of material just outside the shell (for instance, the innermost halo material in the case of the intermediate shell).
In all cases, it is implicitly assumed that no significant acceleration or deceleration has taken place,
which may not be accurate, especially for the inner shell that has been shocked due to the action of the fast wind.
In general, the mass loss is seen to increase monotonically over the last \SI{40000}{yr} from about \SI{2e-6}{\msun.yr^{-1}} to about \SI{3e-5}{\msun.yr^{-1}} in the last few thousand years.
Note that the values shown in Figure~\ref{fig:mass-loss-history} are average values over the respective periods,
so the sudden jump that is apparent \SI{10,000}{yr} ago may have been a more gradual transition.
Also, this method can only estimate mass loss before \SI{1500}{yr} ago, which is the kinematic timescale of the youngest components,
which are the blue knot complexes.
Since that time, the mass-loss rate is expected to have declined significantly
due to the central star transitioning to higher effective temperature and smaller radius,
eventually reaching the present-day value of \(\dot{M} = \SI{9.12e-9}{\msun.yr^{-1}}\) at a wind velocity of \(V = \SI{2150}{km.s^{-1}}\) \citep{Herald:2011a}.
This is indicated on the figure by the gray downward pointing arrow.

@@

A significant fraction of planetary nebulae show deviations even from axisymmetry \citep{Soker:2001a}.
In the cases where these deviations are moderate,
then they may be explained by binary interactions in which the stars are on eccentric orbits,
or the presence of hot or cold spots in the AGB star envelope \citep{Soker:2002b}.
Other cases, where the nebular asymmetries occur only at the outer boundary,
are probably due to interaction with the interstellar medium \citep{Ali:2012a}.
However, a significant minority (\(\sim 10\%\)) of nebulae show such large deviations from bipolar symmetry \citep{Bear:2017a}
that it has been proposed that interacting triple systems are responsible for their shaping \citep{Soker:2004b, Glanz:2020a}.
NGC~6210 falls into this category and has been proposed to be the outcome of either a ``tight binary in a wide orbit'' or a ``tight binary merger'' scenario \citep{Soker:2016b}.
However, given the multitude of evolutionary pathways available in hierarchical triple systems \citep{Toonen:2016a, Toonen:2020a},
other scenarios are also possible.
Sub-arc second imaging observations of young planetary nebulae \citep{Sahai:1998a, Sahai:2011a, Hsia:2014a}
and preplanetary nebulae \citep{Sahai:2007a}
show that multipolarities and asymmetries can appear very early in the evolution of the nebula,
even before the central star is hot enough to emit ionizing photons.

@@

In terms of the GISW model, the halo represents the relatively undisturbed remnant of the AGB wind.
The wind is originally neutral,
but it is flash-ionized when the intermediate shell becomes optically thin to EUV radiation.
Since the propagation of the ionization front through the halo is R-type \citep{Kahn:1954a},
the gas initially experiences no mechanical effect and maintains its original density and velocity,
but since the outflow velocity is now subsonic or transonic, the density gradient becomes a pressure gradient that will begin to accelerate the ionized halo.
The acceleration is stronger at smaller radii, which may explain the difference in velocity between the inner and outer halo.
The time required to produce a velocity change \(\Delta v\) is \(t = H \Delta v / c_{\text{s}}^2\),
where \(H\) is the pressure scale height and \(c_{\text{s}} \approx \SI{12}{km.s^{-1}}\) is the isothermal sound speed.
Assuming a power-law density profile, \(\rho \sim R^{-\alpha}\), yields \(H = R / \alpha\), where \(\alpha \approx 4\) is implied by our observations (Figure~\ref{fig:density-plots}).
The transition between the inner and outer halo is observed to occur around \(R \approx \SI{0.22}{pc}\) (Figure~\ref{fig:halo-components}) and the velocity change is \(\Delta v \approx \SI{8}{km.s^{-1}}\), which together imply a halo ionization age of \(t \approx \SI{2500}{yr}\), albeit with considerable uncertainty.

An alternative way of deriving the halo ionization age is to use the observed size of the halo knots,
assuming that they had a very small size when the halo was neutral and have expanded at the sound speed since the halo was ionized.
The result is \(t \approx \SI{2000}{yr}\), which is very similar to the previous estimate. 
This is consistent with the idea that the intermediate shell first became optically thin when the nebula was roughly half its current age and the central star effective temperature was considerably cooler than it is now (\num{40} to \SI{50}{kK}).

\subsubsection{Asymmetries}
\label{sec:asymmetries}
One of the most obvious qualities of the Turtle Nebula is its chaotic and unstructured appearance,
far from the neat symmetries of many other planetary nebulae.
In part this is due to the shear number of symmetry axes in the nebula
(we have found at least five of them),
but there are also asymmetries \emph{along} many of the axes.
In this section we list the principle ways in which the nebula is lop-sided:
\begin{enumerate}[1.]
\item The inner shell emission in \oiii{} and \Ha{} shows brightness asymmetries in both position and velocity.
  The redshifted half of the shell is roughly twice as bright as the blueshifted half,
  while both halves show spatial asymmetries along axis~A, but in opposite senses:
  redshifted emission is 2.5 times brighter on the north side,
  while blueshifted emission is 1.5 times brighter on the south side.
  Because the inner shell dominates the total flux, this leads to a north--south asymmetry in the nebula as a whole,
  with the peak brightness region lying 2~arcsec north of the central star. 
\item In the \heii{} line, the inner shell has an opposite asymmetry: it is brighter to the south than to the north.
  Such an asymmetry is hard to explain if the \chem{He^{2+}} zone of the nebula is ionization-bounded and the central star emits isotropically.
\item The intermediate shell shows an egg-shaped morphology that is lop-sided along its major axis (axis~B),
  extending roughly 30\% further from the star to the WSW than to the ENE.\@ 
\item The outer lobes are probably the \emph{least} asymmetric component of the entire nebula,
  consisting as they do of balanced point-symmetric red-blue pairs of emission components along the axes A, C, and D (Figure~\ref{fig:outer-lobe-systems}).
  However, there is one significant asymmetry: \oiii{} system \SysP{7},
  which corresponds to the N~knot in \nii{} has no corresponding blue-shifted feature to the S.
\item With the low-ionization knot complexes,
  the principle asymmetry is between the redshifted and blueshifted knots.
  The redshifted knots are brighter by a factor of 2.5,
  slower-moving by a factor of 1.5,
  and further from the star by a factor of 2.
\item In the halo, the only significant asymmetry is in the halo knots,
  which are found only to the N and E of the star. 
\end{enumerate}

What is the origin of the asymmetries?
All mechanisms proposed to date \citet{Soker:2004b, Soker:2016b}
involve accretion disk and jet formation in a triple star system.
This may be via accretion of the AGB wind onto a close binary on a wide orbit around the AGB,
or, if the orbit is closer, via a common envelope or grazing envelope phase,
which may or may not result in the merger of the close binary pair.
The descendant system should be a binary or a triple,
depending on whether a merger occurred,
but there is no direct evidence that the Turtle's central star is a binary.
The fact that the basic parameters of the nebula and its central star can be consistently interpreted in terms of single-star evolution (see \S~\ref{sec:comparison-with-post})
suggests that a triple interaction was not involved in the initial ejection of the envelope that forms the intermediate shell,
although it may be responsible for its shaping.
The existence of two such different

@@

NGC~3242 has been claimed to be shaped by a triple system \citep{Soker:1992a}, although its asymmetries are much subtler than those of the Turtle.
